---
title: "4 2024 second cleaning"
author: "Andrea Portt"
format: html
editor: visual
---

# Second cleaning

Quarto documents 1-3 cleaned Migraine Buddy data, connected migraine data to air pollution via postal code-date, and to weather via lat-long-date.

Now there is further cleaning to do before analysis can begin.

```{r}
library("tidyverse")
library("dplyr")

#install.packages("data.table")
library(data.table)

library("stringr")

```

```{r}
#| echo: false
#| print: false

#read in
  MB <- read.csv("~/2024_04_25_mb_ap.csv")

MB <- read.csv("~/MB_air_pollution/2024_04_25_mb_ap.csv")

setwd("~/MB")

library(readr)
MB <- read_csv("2024_04_25_mb_ap.csv")
View(MB)



```

# Set as .Rdata workspace

```{r}
save(MB, file="~/2024_04_25_data.Rdata")



```

# Read in

.Rdata file is smaller than .csv.

```{r}


load(file="~/2024_04_25_data.Rdata")

```

# Random subset

Picking a random subset of 800 participants to work with in initial stages/for setting up code.

```{r}
#Pick a subset to work with
set.seed(123) # Set seed for reproducibility
radnom_sampledata <- MB_mx_immobile %>%
  distinct(random_uid) %>%
  sample_n(size = 800)


data <- left_join(radnom_sampledata, MB_mx_immobile)

save(data, file="~/2024_12_09_random800.Rdata")

load("~/2024_12_09_random800.Rdata")

```

# Clean columns

Removing extra columns that were artifacts of merging steps.

```{r}
#| echo: false
#| print: false


data <- subset(data, select= -c(X, X.1.x, X.x, X.3, X.2, X.1.y, date.y, new))

data <- data %>% 
  rename(
    date = date.x
    )


```

# Set order

```{r}
#|echo: false
#| print: false
 
#Sort into order for the look of things
data <- data %>% arrange(random_uid, date)

# ORDER (IMPORTANT FOR KEEPING THE TIME SERIES SEQUENCE BY random_uid)
#Setkey fixes  the order into the dataset, rather than simply putting the dataset in order
#Should speed up analysis

data <- as.data.table(data)
setkey(data, random_uid, date)

```

# Gender

The Case Time Series design controls for sex and gender, but we may look at gender as an effect modifier.

Sex is not reported in Migraine Buddy, only gender. There is a lot of missingness.

```{r}
#|echo: false
#| print: false

#Stop and Check 
#Does anyone have more than one gender listed?
#For each uid, count genders
#10,215 uids
MB_uid <- MB %>% 
        group_by(random_uid) %>% 
        summarize()

#20,430 genders listed
MB_gender <- MB %>% 
        group_by(random_uid, gender) %>% 
        summarize()

#Drop rows with NA in gender
#Brings us back to 10,215 observations - so each uid has a single listed gender as well as NAs on non-migraine days
MB_gender <- MB_gender[!is.na(MB_gender$gender),]

```

```{r}
# Fill gender up and down by individual
data <- data %>% 
        group_by(random_uid) %>% 
        fill(gender, .direction = "downup")


#for full dataset
# Fill gender up and down by individual
MB <- MB %>% 
        group_by(random_uid) %>% 
        fill(gender, .direction = "downup")



```

# Age - trim

Age: ranges are from -2 to 114

Look at histogram, there's an obvious pattern - a small group of records "aged" -2 to 4 years old. Also nobody older than 80 until a small group apparently over 100 years old

Values less than 5 were considered missing, as were values greater than 100.

Note: Age is also time-varying, is recorded at time of event in event row. (And is controlled for in CTS, but may be effect modifier)

```{r}

ages <- MB %>% 
        group_by(random_uid) %>% 
        count(age, .drop = TRUE)

ages <- ages[!is.na(ages$age),]

hist(ages$age, breaks  = 117)

min(ages$age)
max(ages$age)

#38 individuals w ages < 5
sum(ages$age <5 )

# 9 individuals with ages >100
sum(ages$age > 100)

#Set ages < 5 and > 100 to NA
data$age <- ifelse (data$age < 5 |
                          data$age > 100 
                          , "NA", data$age)

MB$age <- ifelse (MB$age < 5 |
                          MB$age > 100 
                          , "NA", MB$age)

```

# Day month year columns

```{r}

#Need to fill year month day columns where there's no event

#start by dropping the mostly empty columns by those names
data <- subset(data, select = -c(year,month,day))

#Then re-calculate from "date"
#data$date <- as.character(data$date)

data$date <- as.Date(data$date)

#data$date <- as.Date(data$date, "%Y-%m-%d")

data[c('year', 'month', "day")] <- str_split_fixed(data$date, '-', n = Inf)

data %>%
  tidyr::separate(date, into = c('year', 'month', "day"), sep = "-")

#Relocate
data <- data %>%  relocate(c(year,month, day, dow), .after = date)

```

# Other covariates

Short term critical (according to WHO): temperature, seasonality \[controlled for within btrend\], day of week \[dow covariate\], long term trends - also in btrend, and year-month.

Short term other potential: (WHO) : Holidays, (influenza – no known link to migraine unlike respiratory and cardiac), (Me): Fireworks, Humidity, Barometric pressure

## Ontario holidays

<https://search.r-project.org/CRAN/refmans/lares/html/holidays.html>

```{r}
#Add a column of Ontario holidays by date
#Or Canadian and then add Ontario
#https://www.statutoryholidays.com/2019.php

holidays <- as.Date(c("2019-01-01", "2019-02-18", "2019-04-19", "2019-05-20", "2019-07-01", "2019-08-05", "2019-09-02", "2019-10-14", "2019-12-25", "2019-12-26", 
"2018-01-01", "2018-02-19", "2018-03-30", "2018-05-21", "2018-07-01", "2018-08-06", "2018-09-03", "2018-10-08", "2018-12-25", "2018-12-26",
"2017-01-01", "2017-02-20", "2017-04-14", "2017-05-22", "2017-07-01", "2017-08-07", "2017-09-04", "2017-10-09", "2017-12-25", 
"2017-12-26"))

data$holiday <- as.numeric(data$date %in% holidays)

#Relocate
data <- data %>%  relocate(holiday, .after = dow)

```

## Self-reported migraine

Set up in 1\_\[date\]\_migraine_cleaning.qmd so that if the attack type was recorded as migraine by a user at any point, that user is assumed to be self-reporting migraines at every event.

I made this assumption because attack_type isn't always filled.

Now I need to flag onsets by self-reporters as *migraine* onsets.

```{r}


data$mx_onset <- ifelse (data$self_report == "1" &
                              data$onset == "1"
                          , 1, 0 )

#Relocate
data <- data %>%  relocate(mx_onset, .after = onset)

```

# Save as .Rdata

```{r}

save(data, file="~\\2024_09_20_data.Rdata")  

```

## Read back in

```{r}
#MB dataset 
load("~/2024_09_19_MB.Rdata")  

```

# Seasons

Subset dataset so that I have separate winter (Dec-Jan-Feb) and summer (June-July-August) sets.

```{r}

winter <- data %>% 
        filter(month == c("12", "01", "02"))

summer <- data %>% 
        filter(month == c("06", "07", "08"))


```
